Stage-Splitting a Modal Language

### 2014-08-29 paper/semantics.tex
Front the algo examples section with a bit of explanitory text.
Rest of the edits are small nits when reading.

In addition, there is a stage-\bbone\ type $\fut A$, containing encapsulated
stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
opaquely by stage-\bbone\ code, as they cannot be evaluated until stage \bbtwo.
While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  These are the
only ways in \lang\ to alter the stage of a term; we surround their arguments
with braces in \lang\ syntax to clearly indicate stage boundaries within a
program.

@@

\subsection{Staged Evaluation}

@@

Although the code for {\tt fexp} appears very similar to the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
%
% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
%
%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
%
A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment is a mapping
from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
$v$) to residuals---in our example, we represented this environment with
explicit substitutions. For all of the normal features of \lang
(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
effect on values as standard evaluation, and the final future environment is
gotten by merging the future environments of the subterms.

>>>>

In addition, there is a stage-\bbone\ type $\fut A$, containing encapsulated
stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
opaquely by stage-\bbone\ code, as they cannot be evaluated until stage \bbtwo.
While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  $\next$ and $\prev$ are the
only ways in \lang\ to alter the stage of a term; we surround their arguments
with braces in \lang\ syntax to clearly indicate stage boundaries within a
program.

@@

\subsection{Staged Evaluation}
\label{sec:stagedsemantics}

@@

Although the code for {\tt fexp} appears very similar to the unstaged mathematical
definition above, it is in fact a staged program. The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns stage-\bbtwo\ result. The {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
%
% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
%
%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
%
A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}

% would like some kind of intro here that anticipates all the complexity.
% The dynamics of \lang\ consists of three types of judgements: $\redonesym$ (stage-\bbone reduction), $\redtwosym$ (speculation), and $\reifysym$ (reification).

The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment is a mapping
from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
$v$) to residuals---in our example, we represented this environment with
explicit substitutions. For all of the normal features of \lang
(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
effect on values as standard evaluation, and the final future environment is
gotten by merging the future environments of the subterms.

### 2014-08-29 paper/splitting.tex
Front the algo examples section with a bit of explanitory text.
Rest of the edits are small nits when reading.

The basic idea is that we want to take a \lang\ term which contains interleaved stage \bbone\ and stage \bbtwo\ code
and split it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.
To facilitate communication between the stages, there is a data structure passed between the stages, 
which is an output of the stage \bbone\ term and and input to the stage \bbtwo\ term.

As with the dynamics, the form of the splitting statement depends on the external stage of the term.
We consider splitting for stage \bbtwo\ terms first, as it has a simpler form.  
For example, take \verb|hold{1+2} < 5|.
We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
which are called the {\em precomputation} and {\em residual}\footnote{We also used the term {\em residual} for the result of partial evaluation.
Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
and {\em evaluation-residual}.}, respectively.

>>>>

The stage splitting operation takes as input a \lang\ term which contains interleaved stage \bbone\ and stage \bbtwo\ code
and splits it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.  
To facilitate communication between the stages, there is a data structure passed between the stages, 
which is an output of the stage \bbone\ term and and input to the stage \bbtwo\ term.

As with the dynamics, the form of the splitting statement depends on the external stage of the term.
We consider splitting of stage \bbtwo\ terms first, as it has a simpler form.  
For example, consider \verb|hold{1+2} < 5|.
We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
which are called the {\em precomputation} and {\em residual}\footnote{We also used the term {\em residual} for the result of partial evaluation.
Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
and {\em evaluation-residual}.}, respectively.

### 2014-08-29 paper/techreport.tex
Front the algo examples section with a bit of explanitory text.
Rest of the edits are small nits when reading.

This document is a tech report on ongoing research into a program transformation
technique which we call {\em splitting}.  The splitting operation statically transforms a program written in a modal language into multiple unstaged (monostage) terms.

The idea of splitting has previously been explored as an optimization technique that improves performance in situations where multi-variate functions are invoked many times, with some inputs being invariant across invocations and others being dynamic (varying across invocations). Prior examples of splitting simple languages include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96}, as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  In this document we define the splitting transformation for the two-stage language \lang, a typed lambda calculus featuring products, sums, and isorecursive types.

>>>>

This document is a tech report on ongoing research into a program transformation
technique which we call {\em stage splitting}.  The stage splitting operation statically transforms a program written in a modal language into multiple unstaged (monostage) terms.

The idea of splitting has previously been explored as an optimization technique that improves performance in situations where multi-variate functions are invoked many times, with some inputs being invariant across invocations and others being dynamic (varying across invocations). Prior examples of splitting include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}. In this document we define the splitting transformation for the two-stage language \lang, a typed lambda calculus featuring products, sums, and isorecursive types.
